\documentclass{beamer}
\usetheme{Boadilla}

\newcommand\PRESENTATIONMODE[0]{signal command to allow beamer to work with the header file}

\newcommand\script[1]{}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;} 

\input{../proofs/header.tex}

\title{A Denotational Semantics for Polymorphic Effect Systems}
\subtitle{Part III Project}
\author{Alexander Taylor, at736}
\institute{University of Cambridge}
\date{\today}
\setbeamertemplate{itemize items}[ball]

\newcommand\comment[1]{
    %
}



\comment{
How to give a scientific talk
- Most of the audience cannot refer to thing's you've already said
- Can't ask qs (3-4 qs max)
- therefore, lapses of attention damage comprehension
- therefore talk define about managing attention

    - talk too hard/easy
    - motive unclear

- managing attention with narrative
    - stories + raise further questions
    - only simple narrative needed 
    - for 20 min talk  
        - 6 min motivation
        - 3 min contributions + key ideas
        - 9 mins of explanation
        - 2 mins conclusion

    - motivation
        - Top down: here is a problem, how do we solve it?
            - start with a problem, obvious that a solution is needed to audience
            - motivate search for solution
            - danger 
                - fail to clearly articulate the problem 
                - take too long to explain the problem
        - bottom up: we have observations, which raise questions
            - interesting observations
            - surprising connections between them
            - motivates the search for an explanation

            - dangers
                - boring observatiosn
                - connection boring to non-experts

    The Goal
        - Give your approach to the problem
        - identify one key idea of your solution
            - not zero
                - need to explain to something new
                - doesn't answer problem
            - not n+2
                - many papers require several innovations
                - find a unified way to describe this as one idea
                - otherwise the audience will be confused
    divide and conquer
        - divide the explanation into segment
        - treat each segment as mini talk  
            - motivation, idea, explanantion, conclusion
        - remind audience of what you did and why 
}


\begin{document}
\setbeamertemplate{itemize items}[ball]


\begin{frame}
    \titlepage
\end{frame}

\begin{frame}[fragile]{Motivating Polymorphic Effect Analysis}    

    \todo{Syntax highlight}
\begin{verbatim}
def logAction(
    action: Unit => String
): Unit {
    log.info(action())
}  
\end{verbatim}

\begin{verbatim}
logAction(() => FireMissiles(); "Launched Missiles)

logAction(() => throwError("My Error"))

logAction(() => readEnvironmentVariables)
\end{verbatim}

    \script{
        - Imagine you're a compiler trying to analyse the side effects of this \texttt{doAction} function here.

        - You're not having the best of times
        - it's used three times with three different side effect - one's irrevocable - you can't speculatively execute it, one alters control flow, and the last accesses immutable state so you can do all sorts of optimisations.
        
        - Using some simple  analysis, you can create three separate instantiations of the function to analyse, so it's not too difficult in this case.
        - However in programs with higher order functions, creating these separate instantiations is undecidable in general

        - With a polymorphic effect system, it becomes a bit easier to analyse programs like this. But how do you know that your tools and optimisations are sound?
    }
\end{frame}

\begin{frame}{What is a Denotational Semantics? }
    \begin{itemize}\setlength\itemsep{3em}
        \item A \textit{compositional} mapping $\deno{-}: \text{Language Structure} \rightarrow \text{Mathematical Structure}$
        \item In particular want to define $\deno{\gtyperelation{t}{A}}$
        \item Needs to be \textit{sound} $t_1 \beequiv t_2 \implies \deno{t_1} = \deno{t_2}$
        \item And \textit{adequate} for our needs $\deno{t_1} = \deno{d_2}\implies t_1 \beequiv t_2$
    \end{itemize}

    \script{
        - Denotational semantics is where we create a mapping between program terms and an abstract mathematical structure.
        - We want this to be compositional - meaning that each term's denotation is defined in terms of its subterms 
        - It should also be sound, meaning that equivalent program terms up-to, for example, beta-eta reduction, have equal denotations.
        - we also want it to be adequate - if two terms have equal denotations, then that should mean something useful about the relation of the two terms (e.g. can replace one by the other, etc)
    }
\end{frame}

\begin{frame}
    \frametitle{Contributions}

    \begin{itemize}\setlength\itemsep{3em}
        \item A sound set of requirements for denotational semantics of effect-polymorphic languages.
        \item A method to construct models for effect-polymorphic languages in $\set$.
        \item A proof of adequacy of such a model.
    \end{itemize}

    \script{
        - There are three main contributions here
            - A set of requirements to make a sound model of an effect polymorphic language
            - A construction of models for polymorphic languages in the category of Sets
            - And for a particular language I've proved adequacy.
    }
\end{frame}

\begin{frame}{Denotational Semantics using Category Theory}
    \begin{itemize}
        \setlength\itemsep{3em}
        \item Interested in: Objects, Morphisms, and Functors
        \item $\deno{A}, \deno{\G} \in\obj\C$
        \item $\deno{\gtyperelation{t}{A}}: \deno{\G}\rightarrow \deno{A}$
    \end{itemize}


    \script{
        - In part II, we used domains to handle the semantics of non-termination. Lurking beneath this notion is the idea that we can use category theoretic structure to construct our denotations.

        - Here we map types and type environments to objects in a particular category, and correctly typed terms to morphisms (arrows) in the category.

        - One more structure we need is a functor - a map of objects to objects and morphisms to morphisms that preserves composition of terms.
    }
\end{frame}

\begin{frame}{Language features (1) - Lambda Calculus}

        A cartesian closed category (CCC) consists of:
        \begin{itemize}
            \setlength\itemsep{3em}
            \item Products $A\times B$ - models tuples
            \item A terminal object $\1$ - models the \U\texttt{  }type
            \item Exponential objects $B^A$ - models functions as first-class objects
        \end{itemize}

        \script{
            Now let's look at what category theoretic structure you need to model different language features.
            
            - For lambda calculus style languages, you need at least a cartesian closed category.

            - In terms of the category of sets, this gives you sets of pairs of objects, sets of functions, and the single element set.

            - These are used to model pairs, lambda terms, giving us higher order functions, and the terminal, single element set allows us to interpret the unit type and ground terms.
        }
\end{frame}

\begin{frame}{Language features (2.A) - Monads}
        A (strong) monad consists of:
        \begin{itemize}
            \setlength\itemsep{3em}
            \item A functor $\T{}{}: \C\rightarrow \C$
            \item Join and Unit natural transformations
            \begin{itemize}
                \item  $\bindmu_{A}: \T{}{\T{}{A}}\rightarrow \T{}{A}$
                \item $\point{A}: A\rightarrow \T{}{A}$
            \end{itemize}
            \item Tensor strength natural transformation $\strengtht_{A, B}: A\times\T{}{B}\rightarrow \T{}{(A\times B)}$
        \end{itemize}
        
        \script{
            - To handle effectful computations in a pure language like haskell, you need to use something called a monad.

            - This actually comes from when Moggi first discovered how to model effectful programs.
            - You need this structure of a functor from your category to itself, together with some natural transformation operations for creating and composing effectful operations.
            - If you squint a bit, they even look like the programming language monad definition.

            - An issue with haskell is that each monad only gives you one effect. If you want a program that does IO, contains state, has exceptions, and does non-determinism, you end up with a stack of monads.
            - This is still imprecise for meaningful analysis
        }


\end{frame}

\begin{frame}{Language features (2.B) - Graded Monads}
    A (strong) graded monad consists of:
        \begin{itemize}
            \setlength\itemsep{3em}
            \item An indexed functor $\T{\e}{}: \C\rightarrow \C$
            \item Indexed Join and Unit natural transformations
            \begin{itemize}
                \item  $\bind{\e_1}{\e_2}{A}: \T{\e_1}{\T{\e_2}{A}}\rightarrow \T{\e_1\dot \e_2}{A}$\
                \item $\point{A}: A\rightarrow \T{\1}{A}$
            \end{itemize}
            \item Tensor strength natural transformation $\tstrength{\e}{A}{B}: A\times\T{\e}{B}\rightarrow \T{\e}{(A\times B)}$
        \end{itemize}

        \script{
            - A solution is a graded monad.
            - We now have one functor, indexed by a monoid of effect symbols. The rules and morphisms stay the same but are now also indexed. 

        }    
\end{frame}

\begin{frame}{An Effectful Language}
    \begin{align*}
        v ::= & \const{A} \mid x\mid \t \mid\f \mid\u\mid\lam{x}{A}{v}\mid\apply{v_1}{v_2}\mid\return{v}\\
        & \mid\doin{x}{v_1}{v_2}\mid\pifthenelse{A}{v}{v_1}{v_2} 
    \end{align*}
    
    \begin{align*}
        A, B, C ::= \g\mid \ab \mid\mea
    \end{align*}

    \[
        \ntreeruleI{Return}{\gtyperelation{v}{A}}{\gtyperelation{\return{v}}{\moa}}
    \quad
    \ntreeruleII{Apply}{\gtyperelation{v_1}{\ab}}{\gtyperelation{v_2}{A}}{\gtyperelation{\apply{v_1}{v_2}}{B}}
    \]

    \script{
        - Let's now describe a simple effectful, lambda-calculus-based, language, EC.
        - It has an explicit graded monad, subtyping, if-statements
    }
\end{frame}

\begin{frame}{Semantics of EC}
    \begin{itemize}
        \setlength\itemsep{3em}
        \item Can build a model of EC when we have
        \begin{itemize}
            \item CCC
            \item Strong Graded Monad
            \item Co-product and Subtyping (morphisms for if-statements)
        \end{itemize}
        \item We'll call this an S-category
    \end{itemize}
    
    \[
        \scalebox{.8}{$
        \ntreeruleI{Return}{f = \deno{\gtyperelation{v}{A}}}{\deno{\gtyperelation{\return{v}}{\moa}} = \point{A} \after f}
        \quad
        \ntreeruleI{Fn}{f = \deno{\typerelation{\gax}{v}{B}} : \G \times A \rightarrow B}{\deno{\gtyperelation{\lam{x}{A}{v}}{\ab}} = \cur{f} : \G \rightarrow B^A}
        $}
    \]

    \script{
        - As described, all of the language features can be modelled in a cartesian closed category with a graded monad, a coproduct, and subtyping morphisms.

        - Known as an S-Category

        - Here's an example, if we have the denotation of an expression here, we can get the denotation of using it as a pure computation by postcomposing with the unit of the graded monad.

    }
\end{frame}

\begin{frame}[fragile]{An Ugly Example}
\todo{syntax highlight this}
    \begin{lstlisting}[
        mathescape,
        columns=fullflexible,
        basicstyle=\fontfamily{lmvtt}\selectfont,
      ]
let $twiceIO$ = $\lambda$ $action$: $\M{\texttt{IO}}{\U}$. (
    do $\_$ $\texttt{<-}$ $action$ in $action$
)

let $twiceState$ = $\lambda$ $action$: $\M{\texttt{State}}{\U}$. (
    do $\_$ $\texttt{<-}$ $action$ in $action$
)

do $\_$ $\texttt{<-}$ $twiceState$(increment) in $twiceIO$(writeLog)
    \end{lstlisting}
    \script{
        - Here's a program that's quite ugly with lots of code reuse, since we explicitly state types in the expressions. 
        - we end up with two instantiations of the same function - more work for the compiler
    }
\end{frame}

\begin{frame}{Let's Add Polymorphism}
   
\begin{align*}
    v::=\text{..}\mid\elam{\a}{v}\mid\eapp{v}{\e}
\end{align*}

\begin{align*}
    A, B, C::=\text{...}\mid \all{\a}{A}
\end{align*}

\begin{align*}
    \e ::= e \mid \a \mid \e\dot\e
\end{align*}

    \[
        \ntreeruleI{Effect-Gen}{\etyperelation{\P,\a}{\G}{v}{A}}{\gpetyperelation{\elam{\a}{v}}{\all{\a}{A}}}
        \quad
        \ntreeruleII{Effect-Spec}{\gpetyperelation{v}{\all{\a}A}}{\wellformed{\P}{\e}}{\gpetyperelation{\eapp{v}{\e}}{A\ssub{\a}{\e}}}    
    \]
    \script{
        - Let's now add some polymorphism syntax to EC.
        - This gives us the polymorphic effect calculus.
        - Notice now that in our type rules, we now have an effect-variable environment. Important later
    }
\end{frame}

\begin{frame}[fragile]{An Ugly Example - With a Makeover}

    \todo{Syntax highlighting}
    \begin{lstlisting}[
        mathescape,
        columns=fullflexible,
        basicstyle=\fontfamily{lmvtt}\selectfont,
      ]
let $twice$ = $\Lambda$ $eff$.(
    $\lambda$ $action$: $\M{eff}{\U}$. (
        do $\_$ $\texttt{<-}$ $action$ in $action$
    )
)

do $\_$ $\texttt{<-}$ ($twice$ State increment) in ($twice$ IO writeLog)
    \end{lstlisting}
    \script{
        - We've now added polymorphism, which makes our ugly program more concise.
        - We can reuse this function
    }
\end{frame}

\begin{frame}{How do we Model the Semantics of a Polymorphic Language?}

    \begin{itemize}
        \setlength\itemsep{3em}
        \item For a fixed effect variable environment $\P$ and terms with no polymorphic sub-terms, we have EC
        \item Effect-variable environments of length $n$ are isomorphic by $\a$-equivalence
    \end{itemize}
    
    

    \script{
        - If we fix the effect-variable environment, and disallow polymorphic terms, then PEC terms become EC terms for a particular instantiation.
        - We already know how to build models for PEC
        -It's also the case that effect-variable environments can type the same set of relations, upto alpha equivalence.
        - So there is a countable set of these EC instantiations
    }
\end{frame}

\begin{frame}{How do we Model the Semantics of a Polymorphic Language?}

\begin{minipage}{0.45\linewidth}
    \begin{itemize}
        \item So we instantiate an S-category for each environment.
        \item The type rule for quantification requires us to move between categories \todo{Type rule here}.
        \item Functors are required.
    \end{itemize}
\end{minipage}\quad \begin{minipage}{0.45\linewidth}
    \begin{tikzpicture}[->,>=stealth', scale=0.7, every node/.style={scale=0.7}]
        %draw the s-category stack
\foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {2, 3, 4, 5, 6}{
    \node[ellipse, draw,fill=yellow, fill opacity=1, minimum width=5cm, minimum height=15mm,label=right:$S_\yi$] (s\yi) at (8,\y){};
    \node[circle, draw, inner sep=1pt, label=above:{$\G$}] (g\yi) at (7,\y){};
    \node[circle, draw, inner sep=1pt, fill, label=above:{$A$}] (a\yi) at (9,\y){};
    \draw[->](g\yi) to[bend right=5] node[below]{\tiny $\deno{\etyperelation{\P_\yi}{\G}{v}{A}}$} (a\yi);
}

% Hidden ellipse to draw functors to
\node[ellipse, minimum width=5cm, minimum height=15mm] (s5) at (8,8){};

%Draw the ... for the s-category stack
\foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {7, 7.5, 8}{
    \node[fill, circle, inner sep=1pt] (p\yi) at (8, \y){};
}

% draw the re-indexing functors


%Draw the bracket

\draw [decoration={brace,amplitude=8pt},decorate] ($(s5)+(10em,1ex)$) -- ($(s0)+(10em,-1ex)$);
\node[text width=20mm] (Label) at (14,5){Fibres for each effect environment};
\end{tikzpicture}

\end{minipage}

\script{
    - So we can imagine a stack of these S-Categories, called fibres
    - In order to model polymorphism, we need to have ways of moving morphisms between these fibres - we need functors
}
\end{frame}

\begin{frame}{Base Category}
\begin{itemize}
    \item Need to be able to reason about the effect environment categorically
    \item Model effects and their environments using category $\Eff$: \begin{itemize}
        \item Objects are products of the set of ground effects $\{\singleton\} , E, E\times E,... E^n..$
        \item Morphisms are monotone functions
    \end{itemize}
    \item Represent $\P$ as an object $E^n$
    \item Transformations (e.g. substitutions) between environments become functions
\end{itemize}


    \script{
        - So we need a way of reasoning about the transformations of effect-variable environments in a category theoretic manner
        - To do this, we need a base category, containing a terminal object, an object $U$ representing the kind of effects, and finite products on $U$ 
        - $U^n$ represents the effect-variable environment of length $n$. We'll now use $I$ to indicate this.
        
        - Since all the ground effects have morphisms in this category, we can construct morphisms for substitutions and weakenings of the effect environment.
    }
\end{frame}

\begin{frame}{Indexed Category}

    \begin{tikzpicture}[->,>=stealth', scale=0.7, every node/.style={scale=0.7}]
        % Draw the env objects
        \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {3, 4, 5}{
            \node[fill,circle,inner sep=2pt,label=left:{\small $\P_\yi$}] (d\yi) at (0,\y) {};
        }

        % draw the ... above the env objects
        \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {5.5, 6, 6.5}{
            \node[fill, circle, inner sep=1pt] (dd\yi) at (0,\y){};
        }
        % Draw the index category
        \node[fit=(d0) (d1) (d2) (dd0) (dd1) (dd2),ellipse,draw,minimum width=2cm] {};

        %draw the s-category stack
        \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {2, 4, 6}{
            \node[circle, draw, inner sep=1pt, fill, label=above:{$\G$}] (g\yi) at (7,\y){};
            \node[circle, draw, inner sep=1pt, fill, label=above:{$A$}] (a\yi) at (9,\y){};
            \draw[->](g\yi) to[bend right=5] node[below]{\tiny $\deno{\etyperelation{\P_\yi}{\G}{v}{A}}$} (a\yi);
            \node[ellipse, draw, minimum width=5cm, minimum height=15mm,label=right:$S_\yi$] (s\yi) at (8,\y){};
        }

        % Hidden ellipse to draw functors to
        \node[ellipse, minimum width=5cm, minimum height=15mm] (s3) at (8,8){};

        %Draw the ... for the s-category stack
        \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {7, 7.5, 8}{
            \node[fill, circle, inner sep=1pt] (p\yi) at (8, \y){};
        }

        % Draw index arrows
        \foreach \i in {0, 1, 2}{
            \draw[->, thick] (d\i) to (s\i);
        }

        % draw the re-indexing functors

        \foreach \source[count=\dest] in {0, 1, 2}{
            \draw[->, thick](s\source.north west) to[bend left=10] node[left]{$\pstar$} (s\dest.south west);
        }

        % Draw the quantification functors
        %\foreach \dest[count=\source] in {0, 1, 2}{
        %    \draw[->, thick] 
        %    (s\source.south east) to[bend left=10] node[right]{$\forall_{\P_\dest}$} (s\dest.north east);
        %}

        % Draw the internal morphisms in base category
        \foreach \dest[count=\source] in {0, 1}{
            \draw[->]
            (d\source) to[bend right=10] node[right]{$\p$} (d\dest);
        }

        %Draw the bracket

        \draw [decoration={brace,amplitude=8pt},decorate] ($(s2)+(10em,1ex)$) -- ($(s0)+(10em,-1ex)$);
        \node[text width=20mm] (Label) at (14,5){Fibres for each effect environment};
    \end{tikzpicture}


    \script{
        - So now we can construct this structure
        - called an index category
        - We have a functor mapping each object representing an effect-variable environment to the relevant S-category fibre.
        - Also contravariantly, meaning the direction of the morphism changes, maps morphisms in the base category to re-indexing functors between the relevant fibres.
        - So now we can perform substitutions and weakenings on the effect environment and get the right behaviour on the semantics of the EC instantiation
    }
    
\end{frame}

\begin{frame}{Quantification}
    \begin{itemize}\setlength\itemsep{2em}
        \item What about effect-generalisation?
        \item $\ntreeruleI{Effect-Gen}{\etyperelation{\P,\a}{\G}{v}{A}}{\etyperelation{\P}{\G}{\elam{\a}{v}}{\all{\a}{A}}}$
        \item Need to map $\deno{\etyperelation{\P,\a}{\G}{v}{A}}$ to $\deno{\etyperelation{\P}{\G}{\elam{\a}{v}}{\all{\a}{A}}}$
        \item For specialisation to work, needs: $\pstar\dashv\allI$
    \end{itemize}


    
    \script{
        - We now need to think about the polymorphic terms
        - For quantification, we need another functor, $\allI$ which maps a type rule with an extra effect variable to one quantified over that variable
        - In order for specialisation to work, this quantification functor needs to be a right adjoint to the opposite operation of weakening the effect environment
        - Adjunction of functors is essentially a weaker version of an isomorphism of the categories they're between.
        - Going out on one functor and coming back on the other isn't quite the same as the identity, but it has a well defined action.
    }

\end{frame}
    
\begin{frame}{Instantiating a Model (1)}

        \begin{tikzpicture}[->,>=stealth', scale=0.5, every node/.style={scale=0.5}]
            % Draw the env objects
            \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {3, 4, 5}{
                \node[fill,circle,inner sep=2pt,label=left:{\small $\P_\yi$}] (d\yi) at (0,\y) {};
            }
    
            % draw the ... above the env objects
            \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {5.5, 6, 6.5}{
                \node[fill, circle, inner sep=1pt] (dd\yi) at (0,\y){};
            }
            % Draw the index category
            \node[fit=(d0) (d1) (d2) (dd0) (dd1) (dd2),ellipse,draw,minimum width=2cm] {};
    
            %draw the s-category stack
            \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {2, 4, 6}{
                \node[circle, draw, inner sep=1pt, fill, label=above:{$\G$}] (g\yi) at (7,\y){};
                \node[circle, draw, inner sep=1pt, fill, label=above:{$A$}] (a\yi) at (9,\y){};
                \draw[->](g\yi) to[bend right=5] node[below]{\tiny $\deno{\etyperelation{\P_\yi}{\G}{v}{A}}$} (a\yi);
                \node[ellipse, draw, minimum width=5cm, minimum height=15mm,label=right:$S_\yi$] (s\yi) at (8,\y){};
            }
    
            % Hidden ellipse to draw functors to
            \node[ellipse, minimum width=5cm, minimum height=15mm] (s3) at (8,8){};
    
            %Draw the ... for the s-category stack
            \foreach \y[count=\c,evaluate={\yi=int(\c-1)}] in {7, 7.5, 8}{
                \node[fill, circle, inner sep=1pt] (p\yi) at (8, \y){};
            }
    
            % Draw index arrows
            \foreach \i in {0, 1, 2}{
                \draw[->, thick] (d\i) to (s\i);
            }
    
            % draw the re-indexing functors
    
            \foreach \source[count=\dest] in {0, 1, 2}{
                \draw[->, thick](s\source.north west) to[bend left=10] node[left]{$\pstar$} (s\dest.south west);
            }
    
            % Draw the quantification functors
            \foreach \dest[count=\source] in {0, 1, 2}{
                \draw[->, thick] 
                (s\source.south east) to[bend left=10] node[right]{$\forall_{\P_\dest}$} (s\dest.north east);
            }
    
            % Draw the internal morphisms in base category
            \foreach \dest[count=\source] in {0, 1}{
                \draw[->]
                (d\source) to[bend right=10] node[right]{$\p$} (d\dest);
            }
    
            %Draw the bracket
    
            \draw [decoration={brace,amplitude=8pt},decorate] ($(s2)+(10em,1ex)$) -- ($(s0)+(10em,-1ex)$);
            \node[text width=20mm] (Label) at (14,5){Fibres for each effect environment};
        \end{tikzpicture}

    \begin{itemize}
        \item Can we actually instantiate a category with the required structure? 
        \item Starting point: a model of EC in $\set$
    \end{itemize}
    
    \script{
        - So far, we've only said what structures are required to model PEC.
        - Haven't shown that there actually exists an indexed category with the required properties
        - so let's do that
        - It's fairly well known that you can instantiate models of languages with the same features as EC in the category of sets and functions
        - We want to extend one of these models into one for PEC
    }
\end{frame}

\begin{frame}{Instantiating a Model (2) - Fibres}
    \begin{itemize}\setlength\itemsep{3em}
        \item The fibre $\C(n)$ is the category of functors $[E^n, \set]$
        \item I.E. objects are functions that take a vector of ground effects and return a set $\deno{\wellformedt{\P}{A}}: E^n \rightarrow \obj(\set)$.
        \item Morphisms are dependent functions that return functions in $f: (\ev: E^n)\rightarrow A\ev\rightarrow B\ev$
        \item These fibres have S-Category features
    \end{itemize}

    \script{
        - Now we can think about the fibres
        - These are functor categories
        - I.E. their objects are functions returning sets and their morphisms are dependently typed functions
        - We can construct all the of the S-Category structure  as pointwise
    }
\end{frame}

\begin{frame}{Instantiating a Model (4) - Functors and Adjunctions}
\begin{itemize}
    \item Re-indexing functors act by pre-composition
    \begin{align*}
        A\in&\quad [E^n, \set]\\
        \theta\star(A) \emv =&\quad  A(\theta(\emv))\\
        \theta\star(f) \emv =&\quad f(\theta(\emv)): \theta\star(A) \rightarrow \theta\star(B)\\
    \end{align*}
    \item The quantification functor takes a product over all ground effects
    \begin{align*}
        \allEn(A)\env =&\Pi_{\e\in E}{A(\env, \e)}
    \end{align*}
\end{itemize}
    


    

    \script{
        - Finally, we need to construct the various functors
        - Reindexing functors are done using precomposition
        - quantification consists of a product over all ground effects
    }
\end{frame}

\begin{frame}{The End}
    \begin{itemize}\setlength\itemsep{3em}
        \item Sound: Proved for all indexed S-Categories \checkmark
        \item Compositional: By the definition of denotations \checkmark
        \item Adequate: Proved for an instantiation in $\set$ \checkmark
    \end{itemize}

    \todo{Dissertation and github links}

    \script{Thanks}
    
\end{frame}



\end{document}